---
title: "alpha_div_deltas"
output: html_document
date: "2025-02-26"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(magrittr)
library(broom)
library(qiime2R)
library(apppleplots)
library(ggpubr)
```

**Functions**
`source()` pulls in functions placed in other R scripts and saves them to your global environment!
- here i'm pulling in the statistical analysis functions:
  - `art_lmer_notStrat()` - runs aligned rank transform lmer on the data in an overall model (not stratified)
  - `art_lmer_strat()` - runs aligned rank transform lmer on the data in a variable stratified model
  - `art_lmer_doubleStrat()` - runs aligned rank transform lmer on the data in a model stratified by two variables
```{r}
source("art_lmer_functions.R")
```

**File paths**
```{r}
faith_pd_fp <- '../../oren_crestone/data/qiime/core_outputs/faith_pd.tsv'
shannon_fp <- '../../oren_crestone/data/qiime/core_outputs/shannon_entropy.tsv'
proc_seq_meta_fp <- '../data/proc_seq_metadata.tsv'
subject_info_fp <- '../data/subject_info.tsv'

treat_group_labs <- c('CRS3123 200mg',
                      'CRS3123 400mg',
                      'Vancomycin 125mg')
```

**Reading in files**
```{r}
faith_pd <- read_tsv(faith_pd_fp) %>% 
  rename(sampleid = `#SampleID`)

shannon <- read_tsv(shannon_fp) %>% 
  rename(sampleid = `...1`)

proc_seq_meta <- read_tsv(proc_seq_meta_fp)

subject_info_meta <- read_tsv(subject_info_fp)
```

**Data wrangled**
```{r}
## combining faith, shannon, and clinical meta w seq metadata
alphaDiv_seq_meta <- proc_seq_meta %>% 
  left_join(faith_pd, by = 'sampleid') %>% 
  left_join(shannon, by = 'sampleid') %>% 
  left_join(subject_info_meta, by = 'subject_id') %>% 
  na.omit()

## calculating faith's pd deltas between visits 2 and 4 (more samples from visit 2)
d1_toc_deltas <- alphaDiv_seq_meta %>% 
  select(subject_id, treatment_group, new_visit_label, faith_pd) %>% 
  filter(new_visit_label == 'D1' | new_visit_label == 'D10-15/TOC') %>% 
  spread(new_visit_label, faith_pd) %>% 
  na.omit() %>% 
  group_by(subject_id) %>% 
  mutate('D1_TOC_diff' = (`D10-15/TOC` - D1)) %>% 
  convert_as_factor(subject_id, treatment_group)

## calculating faith's pd deltas between visits 1 and 4 
screen_toc_deltas <- alphaDiv_seq_meta %>% 
  select(subject_id, treatment_group, new_visit_label, faith_pd) %>% 
  filter(new_visit_label == 'Screen' | new_visit_label == 'D10-15/TOC') %>% 
  spread(new_visit_label, faith_pd) %>% 
  na.omit() %>% 
  group_by(subject_id) %>% 
  mutate('Screen_TOC_diff' = (`D10-15/TOC` - Screen)) %>% 
  convert_as_factor(subject_id, treatment_group)

## calculating faith's pd deltas between visits 4 and 7
toc_d70_deltas <- alphaDiv_seq_meta %>% 
  select(subject_id, treatment_group, new_visit_label, faith_pd) %>% 
  filter(new_visit_label == 'D10-15/TOC' | new_visit_label == 'D70') %>% 
  spread(new_visit_label, faith_pd) %>% 
  na.omit() %>% 
  group_by(subject_id) %>% 
  mutate('D70_TOC_diff' = (D70 - `D10-15/TOC`)) %>% 
  convert_as_factor(subject_id, treatment_group)

## calculating faith's pd deltas between visits 4 and 6 (more samples from visit 6)
toc_d40_deltas <- alphaDiv_seq_meta %>% 
  select(subject_id, treatment_group, new_visit_label, faith_pd) %>% 
  filter(new_visit_label == 'D10-15/TOC' | new_visit_label == 'D40') %>% 
  spread(new_visit_label, faith_pd) %>% 
  na.omit() %>% 
  group_by(subject_id) %>% 
  mutate('D40_TOC_diff' = (D40 - `D10-15/TOC`)) %>% 
  convert_as_factor(subject_id, treatment_group)

```

**Stats** 
creating a named list of delta dataframes and a list of what their names are
```{r}
delta_df_list <- list(Screen_TOC_diff = screen_toc_deltas,
                      D1_TOC_diff = d1_toc_deltas,
                      D40_TOC_diff = toc_d40_deltas,
                      D70_TOC_diff = toc_d70_deltas)

delta_df_names <- c('Screen_TOC_diff',
                    'D1_TOC_diff',
                    'D40_TOC_diff',
                    'D70_TOC_diff')
```

using a for loop to extract each deltas dataframe out of the named list, run stats on it, add a column of which dataframe its from for the results and combining all the resulting stats dfs together (could i have made this easier on myself? sure but i didn't want to)
```{r}
out_res <- tibble()
out_contrast <- tibble()

for (df_name in unique(unlist(delta_df_names))){
  wanted_df <- delta_df_list[[df_name]]
  
  deltas_lmer <- art_lmer_notStrat(input_table = wanted_df,
                                   formula_left = df_name,
                                   formula_right = 'treatment_group',
                                   contrast_run_by = 'treatment_group')

  deltas_lmer_res <- deltas_lmer$ARTResults %>% 
    mutate(deltas_calc_between = paste(df_name)) %>% 
    select(deltas_calc_between, everything())
  
  deltas_lmer_contrast <- deltas_lmer$ARTContrast %>% 
    mutate(deltas_calc_between = paste(df_name)) %>% 
    select(deltas_calc_between, everything())
  
  out_res <- bind_rows(out_res, deltas_lmer_res)
  out_contrast <- bind_rows(out_contrast, deltas_lmer_contrast)
}
```

**D1-TOC Faith's PD Plot**
```{r, fig.width=10, fig.height=8}
d1_toc_contrast <- out_contrast %>% 
  filter(deltas_calc_between == 'D1_TOC_diff')

(d1_toc_deltas_plot <- d1_toc_deltas %>% 
  ggplot(aes(x = treatment_group, y = D1_TOC_diff)) +
  geom_boxplot(aes(group = treatment_group), outlier.shape = NA) +
  geom_hline(yintercept = 0, linetype = 'dashed', color = 'purple', size = 0.4) +
  geom_jitter(width = 0.1, height = 0, alpha = 0.4, size = 2.5) +
  theme_bw(base_size = 20) +
  stat_pvalue_manual(d1_toc_contrast,
                     y.position = 25,
                     step.increase = 0.1,
                     label = 'signif') +
  scale_x_discrete(labels = treat_group_labs) +
  labs(x = 'Treatment Group',
       y = "Faith's PD Difference (TOC-D1)",
       title = "Faith's PD: TOC vs D1"))
```


**Saving my outputs**
```{r}
ggsave('../plots/deltas_faithPD_TOC-D1.pdf',
       plot = d1_toc_deltas_plot,
       width = 10,
       height = 8)


write_tsv(out_res,
          '../stats/deltas_faithPD_by_treatGroup.tsv')
write_tsv(out_contrast,
          '../stats/deltas_faithPD_by_treatGroup_postHoc.tsv')
```

