---
title: "taxonomy_stats"
output: html_document
date: "2025-02-25"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(magrittr)
library(broom)
library(qiime2R)
library(apppleplots)
library(ggpubr)
library(rstatix)
```

## **Functions**
`source()` pulls in functions placed in other R scripts and saves them to your global environment!
- here i'm pulling in the statistical analysis functions:
  - `test_normality()` - does basic R tests for normality on the data
  - `art_lmer_notStrat()` - runs aligned rank transform lmer on the data in an overall model (not stratified)
  - `art_lmer_strat()` - runs aligned rank transform lmer on the data in a variable stratified model
  - `art_lmer_doubleStrat()` - runs aligned rank transform lmer on the data in a model stratified by two variables
```{r}
source("art_lmer_functions.R")
source("normality_function.R")

abun_file_prep <- function(metadata_fp,
                           tax_fp,
                           otu_table_fp,
                           group_by_cols,
                           filter_tax_by = NULL,
                           filter_tax_col = NULL){
  ## metadata
  metadata_file <- read_tsv(metadata_fp)
  ## taxonomy
  taxonomy <- read_qza(tax_fp)$data %>% 
    parse_taxonomy() %>% 
    rownames_to_column('asv')
  ## otu table 
  otu_table <- read_qza(otu_table_fp)$data
  
  otu_table <- otu_table %>% 
    as_tibble(rownames = 'asv') %>% 
    gather(-asv, key = sampleid, value = abun) %>% 
    group_by(sampleid) %>% 
    mutate(rel_abun = abun/sum(abun)) %>% 
    mutate(rel_abun = rel_abun + 0.000001) %>% 
    filter(sampleid %in% metadata_file$sampleid)
  ## joining all tables together 
  abun_table <- otu_table %>% 
    left_join(metadata_file, by = 'sampleid') %>% 
    left_join(taxonomy, by = 'asv') %>% 
    group_by(across(all_of(group_by_cols))) %>%
    summarise(rel_abund = sum(rel_abun)) %>% 
    ungroup()
  
  if (is.character(filter_tax_by) & is.character(filter_tax_col)){
    abun_table <- abun_table %>% 
      filter(.data[[filter_tax_col]] %in% filter_tax_by)
  } else {
    abun_table
  }
  
  ## creating a list for my outputs
  my_list <- list(Taxonomy = taxonomy,
                  OTUTable = otu_table,
                  AbundanceTable = abun_table)
  return(my_list)
}

## small function to assign a value to a key in a dataframe column (i.e. numbers to actual categorical variables)
values_to_keys <- function(input_table,
                           key_col_name,
                           value_col_name,
                           key_value_list,
                           col_order_list){
  
  input_table[value_col_name] <- unlist(lapply(unlist(input_table[key_col_name]), 
                                        FUN = function(x){key_value_list[[x]]}))
  
  output_table <- input_table %>%
    select(all_of(col_order_list), everything())
  
  return(output_table)
}
```

## **Reading in files and data wrangling**

**File paths**
```{r}
tax_FP <- '../../oren_crestone/data/qiime/taxonomy.qza'
otu_table_FP <- '../../oren_crestone/data/qiime/taxOnly_otu_table.qza'
proc_seq_meta_fp <- '../data/proc_seq_metadata.tsv'
subject_info_fp <- '../data/subject_info.tsv'

## global lists for renaming plot labels
visit_x_labs <- c('Screen',
                  'D1',
                  'D4',
                  'TOC',
                  'D17', 
                  'D40', 
                  'D70') 

treat_group_labs <- c('CRS3123 200mg',
                      'CRS3123 400mg',
                      'Vancomycin 125mg')
names(treat_group_labs) <- c('CRS3123_200mg_po_bid',
                             'CRS3123_400mg_po_bid',
                             'Vancomycin_125mg_po_qid')

special_treatGroup_x_labs <- c('CRS3123\n200mg',
                               'CRS3123\n400mg',
                               'Vancomycin\n125mg')
```

**Prepping abundance table**
```{r}
subject_info_df <- read_tsv(subject_info_fp)

abun_col_list <- c('sampleid',
                   'subject_id',
                   'visit_day',
                   'new_visit_num',
                   'new_visit_label',
                   'Phylum',
                   'Family',
                   'Genus')

wanted_family <- c('Enterobacteriaceae')

wanted_genus <- c('Bifidobacterium',
                  'Enterococcus',
                  'Clostridioides',
                  'Lactobacillus',
                  'Bacteroides',
                  'Escherichia-Shigella')

abun_files <- abun_file_prep(metadata_fp = proc_seq_meta_fp,
                             tax_fp = tax_FP,
                             otu_table_fp = otu_table_FP,
                             group_by_cols = abun_col_list,
                             filter_tax_by = wanted_genus,
                             filter_tax_col = 'Genus')

proc_abun_table <- abun_files$AbundanceTable

proc_abun_table <- proc_abun_table %>% 
  left_join(subject_info_df, by = 'subject_id') %>% 
  na.omit() %>% 
  mutate(new_visit_num = as.numeric(new_visit_num))
```
## **Relative abundance over time plot**
```{r, fig.width=17, fig.height=13}
(genusAbun_plot <- apppleplots::facet_twice_plots(input_table = proc_abun_table,
                               x_value = 'new_visit_num',
                               y_value = 'rel_abund',
                               x_value_type = 'discrete',
                               y_transform = TRUE,
                               x_labels = visit_x_labs,
                               box_group_by = 'new_visit_num',
                               line_group_by = 'subject_id',
                               point_alpha = 0,
                               facet_rows = 'Genus',
                               facet_cols = 'treatment_group',
                               row_labs = NULL,
                               col_labs = treat_group_labs,
                               x_name = 'Study Day',
                               y_name = 'Relative Abundance (log10)',
                               title_content = 'Select Genera Relative Abundance') +
  scale_x_continuous(breaks = c(1, 2, 3, 4, 5, 6, 7),
                     labels = visit_x_labs) +
  geom_jitter(width = 0.1, height = 0, alpha = 0.4, size = 2.5) +
  theme_bw(base_size = 20) +
  theme(strip.text.y.right = element_text(angle = 0)))
```

## **ART LMER Stats**

**Not stratified analysis**
```{r}
factored_abun_table <- proc_abun_table %>% 
  select(sampleid, subject_id, new_visit_num, Genus, rel_abund, treatment_group) %>% 
  convert_as_factor(subject_id, new_visit_num, Genus, treatment_group)

genusAbun_art_lmer <- art_lmer_notStrat(input_table = factored_abun_table,
                                        formula_left = 'rel_abund',
                                        formula_right = 'new_visit_num * Genus * treatment_group + (1|subject_id)',
                                        contrast_run_by = 'new_visit_num:Genus:treatment_group')

(genusAbun_notStrat_res <- genusAbun_art_lmer$ARTResults)
(genusAbun_notStrat_contrast <- genusAbun_art_lmer$ARTContrast)
```

**Stratified by treatment group**
```{r}
genusAbun_art_lmer_strat <- art_lmer_strat(input_table = factored_abun_table,
                                           strat_variable = 'treatment_group',
                                           model_formula_left = 'rel_abund',
                                           model_formula_right = 'new_visit_num * Genus + (1|subject_id)',
                                           model_contrast_run_by = 'new_visit_num:Genus')

(genusAbun_strat_res <- genusAbun_art_lmer_strat$ARTResults)
(genusAbun_strat_contrast <- genusAbun_art_lmer_strat$ARTContrast)
```

**Double stratified by genus and visit**
dont need to add subject_id as a random effect to this model since I'm running the stats between relative abundance of the genera individually for each day and treatment group (so you only have one sample per subject bc its not across the whole timeline)
```{r}
genusAbun_art_lmer_doubleStrat <- art_lmer_doubleStrat(input_table = factored_abun_table,
                                                       strat_variable1 = 'Genus',
                                                       strat_variable2 = 'new_visit_num',
                                                       model_formula_left = 'rel_abund',
                                                       model_formula_right = 'treatment_group',
                                                       model_contrast_run_by = 'treatment_group')

(genusAbun_doubleStrat_res <- genusAbun_art_lmer_doubleStrat$ARTResults)
(genusAbun_doubleStrat_contrast <- genusAbun_art_lmer_doubleStrat$ARTContrast)
```

creating a column in the stat results that translates visit number to study day for ease of understanding
```{r}
visit_key_value <- c('1' = 'Screen',
                     '2' = 'D1',
                     '3' = 'D4',
                     '4' = 'D10-15/TOC',
                     '5' = 'D17',
                     '6' = 'D40',
                     '7' = 'D70')
visit_col_order <- c('Genus', 
                     'new_visit_num', 
                     'visit_label')

genusAbun_doubleStrat_res_wValues <- values_to_keys(input_table = genusAbun_doubleStrat_res,
                                                    key_col_name = 'new_visit_num',
                                                    value_col_name = 'visit_label',
                                                    key_value_list = visit_key_value,
                                                    col_order_list = visit_col_order)

genusAbun_doubleStrat_contrast_wValues <- values_to_keys(input_table = genusAbun_doubleStrat_contrast,
                                                         key_col_name = 'new_visit_num',
                                                         value_col_name = 'visit_label',
                                                         key_value_list = visit_key_value,
                                                         col_order_list = visit_col_order)
```


## **Saving my outputs**
```{r}
ggsave('../plots/genus_relAbun_plot.pdf',
       plot = genusAbun_plot,
       width = 17,
       height = 13)

## stats
write_tsv(genusAbun_strat_res,
          '../stats/genusAbun_treatGroup_stratOnly.tsv')
write_tsv(genusAbun_strat_contrast,
          '../stats/genusAbun_treatGroup_stratOnly_postHoc.tsv')
write_tsv(genusAbun_doubleStrat_res_wValues,
          '../stats/genusAbun_treatGroup_visit_strat.tsv')
write_tsv(genusAbun_doubleStrat_contrast_wValues,
          '../stats/genusAbun_treatGroup_visitStrat_postHoc.tsv')
```

