---
title: "taxonomy_deltas"
output: html_document
date: "2025-03-19"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(magrittr)
library(broom)
library(qiime2R)
library(apppleplots)
library(ggpubr)
library(rstatix)
```

## **Functions**
`source()` pulls in functions placed in other R scripts and saves them to your global environment!
- here i'm pulling in the statistical analysis functions:
  - `test_normality()` - does basic R tests for normality on the data
  - `art_lmer_notStrat()` - runs aligned rank transform lmer on the data in an overall model (not stratified)
  - `art_lmer_strat()` - runs aligned rank transform lmer on the data in a variable stratified model
  - `art_lmer_doubleStrat()` - runs aligned rank transform lmer on the data in a model stratified by two variables
```{r}
source("art_lmer_functions.R")
source("normality_function.R")

abun_file_prep <- function(metadata_fp,
                           tax_fp,
                           otu_table_fp,
                           group_by_cols,
                           filter_tax_by = NULL,
                           filter_tax_col = NULL){
  ## metadata
  metadata_file <- read_tsv(metadata_fp)
  ## taxonomy
  taxonomy <- read_qza(tax_fp)$data %>% 
    parse_taxonomy() %>% 
    rownames_to_column('asv')
  ## otu table 
  otu_table <- read_qza(otu_table_fp)$data
  
  otu_table <- otu_table %>% 
    as_tibble(rownames = 'asv') %>% 
    gather(-asv, key = sampleid, value = abun) %>% 
    group_by(sampleid) %>% 
    mutate(rel_abun = abun/sum(abun)) %>% 
    mutate(rel_abun = rel_abun + 0.000001) %>% 
    filter(sampleid %in% metadata_file$sampleid)
  ## joining all tables together 
  abun_table <- otu_table %>% 
    left_join(metadata_file, by = 'sampleid') %>% 
    left_join(taxonomy, by = 'asv') %>% 
    group_by(across(all_of(group_by_cols))) %>%
    summarise(rel_abund = sum(rel_abun)) %>% 
    ungroup()
  
  if (is.character(filter_tax_by) & is.character(filter_tax_col)){
    abun_table <- abun_table %>% 
      filter(.data[[filter_tax_col]] %in% filter_tax_by)
  } else {
    abun_table
  }
  
  ## creating a list for my outputs
  my_list <- list(Taxonomy = taxonomy,
                  OTUTable = otu_table,
                  AbundanceTable = abun_table)
  return(my_list)
}

## function to generate the deltas plot, can be faceted or not 
## stat table must be provided with group 1, group 2, p.adj.signif, and y.position columns 
## (and a column matching what you're faceting by if relevant)
make_deltas_plot <- function(input_table,
                             x_axis,
                             y_axis,
                             facet_by = NULL,
                             free_y = FALSE,
                             stat_table,
                             x_labs,
                             x_name,
                             y_name,
                             title_content){
  plot <- input_table %>% 
    ggplot(aes(x = .data[[x_axis]], y = .data[[y_axis]])) +
    geom_boxplot(aes(group = .data[[x_axis]]), outlier.shape = NA) +
    geom_hline(yintercept = 0, linetype = 'dashed', color = 'purple', size = 0.4) +
    geom_jitter(width = 0.1, height = 0, alpha = 0.4, size = 2.5) +
    theme_bw(base_size = 20) +
    scale_x_discrete(labels = x_labs) +
    labs(x = x_name,
         y = y_name,
         title = title_content)
  
  if (is.character(facet_by) & free_y == TRUE){
    plot <- plot +
       facet_wrap(~.data[[facet_by]],
                  scales = 'free_y') +
       stat_pvalue_manual(stat_table,
                          step.group.by = facet_by,
                          step.increase = 0.1,
                          label = 'p.adj.signif')
  } else {
    if (is.character(facet_by)){
      plot <- plot + 
        facet_wrap(.data[[facet_by]],
                   scales = "fixed") +
        stat_pvalue_manual(stat_table,
                           step.group.by = facet_by,
                           step.increase = 0.1,
                           label = 'p.adj.signif')
    } else {
      plot <- plot +
        stat_pvalue_manual(stat_table,
                           step.increase = 0.1,
                           label = 'p.adj.signif')
    }
  }
  
  return(plot)
}
```


## **Reading in files and data wrangling**

**File paths**
```{r}
tax_FP <- '../../oren_crestone/data/qiime/taxonomy.qza'
otu_table_FP <- '../../oren_crestone/data/qiime/taxOnly_otu_table.qza'
proc_seq_meta_fp <- '../data/proc_seq_metadata.tsv'
subject_info_fp <- '../data/subject_info.tsv'

## global lists for renaming plot labels
visit_x_labs <- c('Screen',
                  'D1',
                  'D4',
                  'TOC',
                  'D17', 
                  'D40', 
                  'D70') 

treat_group_labs <- c('CRS3123 200mg',
                      'CRS3123 400mg',
                      'Vancomycin 125mg')
names(treat_group_labs) <- c('CRS3123_200mg_po_bid',
                             'CRS3123_400mg_po_bid',
                             'Vancomycin_125mg_po_qid')

special_treatGroup_x_labs <- c('CRS3123\n200mg',
                               'CRS3123\n400mg',
                               'Vancomycin\n125mg')
```

**Prepping abundance table**
```{r}
subject_info_df <- read_tsv(subject_info_fp)

abun_col_list <- c('sampleid',
                   'subject_id',
                   'visit_day',
                   'new_visit_num',
                   'new_visit_label',
                   'Phylum',
                   'Family',
                   'Genus')

wanted_family <- c('Enterobacteriaceae')

wanted_genus <- c('Bifidobacterium',
                  'Enterococcus',
                  'Clostridioides',
                  'Lactobacillus',
                  'Bacteroides',
                  'Escherichia-Shigella')

abun_files <- abun_file_prep(metadata_fp = proc_seq_meta_fp,
                             tax_fp = tax_FP,
                             otu_table_fp = otu_table_FP,
                             group_by_cols = abun_col_list,
                             filter_tax_by = wanted_genus,
                             filter_tax_col = 'Genus')

proc_abun_table <- abun_files$AbundanceTable

proc_abun_table <- proc_abun_table %>% 
  left_join(subject_info_df, by = 'subject_id') %>% 
  na.omit() %>% 
  mutate(new_visit_num = as.numeric(new_visit_num))
```


## **Baseline vs TOC rel abun deltas plot for all selected genera**

**Calculating TOC-baseline deltas**
```{r}
## putting together deltas table
all_genera_deltas_table <- proc_abun_table %>% 
  select(subject_id, new_visit_label, Genus, rel_abund, treatment_group) %>% 
  filter(new_visit_label == 'D1' | new_visit_label == 'D10-15/TOC') %>% 
  spread(new_visit_label, rel_abund) %>% 
  na.omit() %>% 
  group_by(subject_id) %>% 
  mutate(D1_TOC_diff = (`D10-15/TOC` - D1)) %>% 
  convert_as_factor(treatment_group)
```

**LMER stats for deltas (stratified by genus)**
```{r}
allGenera_deltas_lmer <- art_lmer_strat(input_table = all_genera_deltas_table,
                                           strat_variable = 'Genus',
                                           model_formula_left = 'D1_TOC_diff',
                                           model_formula_right = 'treatment_group',
                                           model_contrast_run_by = 'treatment_group')

## adding y position to stat table so I can put it on the plot 
allGenera_deltas_contrast <- allGenera_deltas_lmer$ARTContrast %>% 
  rename(p.adj.signif = signif,
         p = p.value) %>%
  mutate(y.position = case_when(
    Genus == 'Bacteroides' ~ 0.7,
    Genus == 'Bifidobacterium' ~ 0.4,
    Genus == 'Clostridioides' ~ 0.04,
    Genus == 'Enterococcus' ~ 0.5,
    Genus == 'Escherichia-Shigella' ~ 0.1,
    Genus == 'Lactobacillus' ~ 0.35
  ))
```

**Plot with stats on it**
```{r, fig.width=15, fig.height=9}
## deltas plot with stats
(allGenera_deltasAbun_plot <- make_deltas_plot(input_table = all_genera_deltas_table,
                                               x_axis = 'treatment_group',
                                               y_axis = 'D1_TOC_diff', 
                                               facet_by = 'Genus',
                                               free_y = TRUE,
                                               stat_table = allGenera_deltas_contrast,
                                               x_labs = special_treatGroup_x_labs,
                                               x_name = 'Treatment Group',
                                               y_name = 'Relative Abundance Difference (TOC-D1)',
                                               title_content = 'Relative Abundance: Baseline vs TOC'))
```


## **TOC vs D40 rel abun deltas plot for all selected genera**

**Calculating D40-TOC deltas**
```{r}
## putting together deltas table
allGenera_deltas_TOC_D40 <- proc_abun_table %>% 
  select(subject_id, new_visit_label, Genus, rel_abund, treatment_group) %>% 
  filter(new_visit_label == 'D40' | new_visit_label == 'D10-15/TOC') %>% 
  spread(new_visit_label, rel_abund) %>% 
  na.omit() %>% 
  group_by(subject_id) %>% 
  mutate(D40_TOC_diff = (D40 - `D10-15/TOC`)) %>% 
  convert_as_factor(treatment_group)
```

**LMER stats for deltas (stratified by genus)**
```{r}
allGenera_deltas_TOC_D40_lmer <- art_lmer_strat(input_table = allGenera_deltas_TOC_D40,
                                           strat_variable = 'Genus',
                                           model_formula_left = 'D40_TOC_diff',
                                           model_formula_right = 'treatment_group',
                                           model_contrast_run_by = 'treatment_group')

## adding y position to stat table so I can put it on the plot 
allGenera_deltas_TOC_D40_contrast <- allGenera_deltas_TOC_D40_lmer$ARTContrast %>% 
  rename(p.adj.signif = signif,
         p = p.value) %>%
  mutate(y.position = case_when(
    Genus == 'Bacteroides' ~ 0.4,
    Genus == 'Bifidobacterium' ~ 0.1,
    Genus == 'Clostridioides' ~ 0.04,
    Genus == 'Enterococcus' ~ 0.1,
    Genus == 'Escherichia-Shigella' ~ 0.3,
    Genus == 'Lactobacillus' ~ 0.1
  ))
```

**Plot with stats**
```{r, fig.width=15, fig.height=9}
## deltas plot with stats
(allGenera_deltasAbun_TOC_D40_plot <- make_deltas_plot(input_table = allGenera_deltas_TOC_D40,
                                                       x_axis = 'treatment_group',
                                                       y_axis = 'D40_TOC_diff', 
                                                       facet_by = 'Genus',
                                                       free_y = TRUE,
                                                       stat_table = allGenera_deltas_TOC_D40_contrast,
                                                       x_labs = special_treatGroup_x_labs,
                                                       x_name = 'Treatment Group',
                                                       y_name = 'Relative Abundance Difference (D40-TOC)',
                                                       title_content = 'Relative Abundance: TOC vs D40'))
```


## **Bifidobacterium deltas: baseline vs TOC**

**Calculating TOC-baseline deltas**
```{r}
bifido_only_table <- proc_abun_table %>% 
  filter(Genus == 'Bifidobacterium') %>% 
  select(subject_id, new_visit_label, Genus, rel_abund, treatment_group) %>% 
  filter(new_visit_label == 'D1' | new_visit_label == 'D10-15/TOC') %>% 
  spread(new_visit_label, rel_abund) %>% 
  na.omit() %>% 
  group_by(subject_id) %>% 
  mutate(D1_TOC_diff = (`D10-15/TOC` - D1)) %>% 
  convert_as_factor(treatment_group)
```

**LMER stats for deltas (don't need to be stratified by anything since just for one genus)**
```{r}
bifido_deltas_lmer <- art_lmer_notStrat(input_table = bifido_only_table,
                                        formula_left = 'D1_TOC_diff',
                                        formula_right = 'treatment_group',
                                        contrast_run_by = 'treatment_group')

## adding y position to stat table so I can put it on the plot 
proc_bifido_deltas_lmer <- bifido_deltas_lmer$ARTContrast %>% 
  rename(p.adj.signif = signif,
         p = p.value) %>% 
  mutate(y.position = as.numeric(paste(0.40)))
```

**Plot with stats**
```{r, fig.width=10, fig.height=8}
## deltas plot with stats
(bifido_deltasAbun_plot <- make_deltas_plot(input_table = bifido_only_table,
                                            x_axis = 'treatment_group',
                                            y_axis = 'D1_TOC_diff',
                                            facet_by = NULL,
                                            free_y = FALSE,
                                            stat_table = proc_bifido_deltas_lmer,
                                            x_labs = treat_group_labs,
                                            x_name = 'Treatment Group',
                                            y_name = 'Relative Abundance Difference (TOC-D1)',
                                            title_content = 'Bifidobacterium Relative Abundance: Baseline vs TOC'))
```


## **Saving my outputs**
```{r}
## plots
ggsave('../plots/relAbun_genus_deltas_TOC-D1.pdf',
       plot = allGenera_deltasAbun_plot,
       width = 15, 
       height = 9)
ggsave('../plots/relAbun_genus_deltas_D40-TOC.pdf',
       plot = allGenera_deltasAbun_TOC_D40_plot,
       width = 15, 
       height = 9)
ggsave('../plots/bifido_abun_deltas.pdf',
       plot = bifido_deltasAbun_plot,
       width = 10,
       height = 8)

## stats
write_tsv(proc_bifido_deltas_lmer,
          '../stats/bifido_abunDeltas_TOC-D1.tsv')
```

